Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Being more precise about properties vs predicates and invariants

\newglossaryentry{atomicity}
{
  name=atomicity,
  description={describes that a certain machine instruction or sequence of machine
    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}
}
\newglossaryentry{concurrent execution}
{
  name=concurrent execution,
  description={is when there are multiple processes executing and their machine
    instructions are interleaved in an unpredictable manner}
}
\newglossaryentry{determinism}
{
  name=determinism,
  description={is when the outcome of an execution is uniquely determined by the
    initial state}
}
\newglossaryentry{invariant}
{
  name=invariant,
  description={is a binary property that must hold for every state of a process}
}
\newglossaryentry{machine instruction}
{
  name=machine instruction,
  description={is an atomic operation on the CXL machine, executed by a process}
}
\newglossaryentry{macro step}
{
  name=macro step,
  description={is a sequence of micro steps with at most one instruction
        externally visible to other processes.  It is the unit of interleaving
        in CXL}
}
\newglossaryentry{micro step}
{
  name=micro step,
  description={is the execution of one machine instruction by a process, updating
        its state and the state of the CXL machine}
}
\newglossaryentry{model checking}
{
  name=model checking,
  description={is a formal verification method that explores all possible executions
    of a program, which must have a finite number of states}
}
\newglossaryentry{process}
{
  name=process,
  description={is a method in execution.  We do not make the distinction
      between processes and threads.  A process has a program counter, a register,
      and a stack}
}
\newglossaryentry{process variable}
{
  name=process variable,
  description={is a variable that is private to a single process and stored in
        its register}
}
\newglossaryentry{sequential execution}
{
  name=sequential execution,
  description={is when there is just one process executing, as opposed to concurrent
    execution}
}
\newglossaryentry{shared variable}
{
  name=shared variable,
  description={is a variable that is stored in the memory of the CXL machine and
      shared between multiple processes, as opposed to a process variable}
}

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
The CXL machine is predominantly a \emph{stack machine}.
\index{stack machine}
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process (with name tag \texttt{\_\_init\_\_/()}
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}:
\index{critical section}
\index{critical region}
a set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) is called \emph{thread-safe}.
\index{thread safety}
\index{thread}

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/states-crop.pdf}
\end{center}
\caption{Venn diagram of all states and a trace.}
\label{fig:states}
\end{figure}

@@

A \emph{property}
\index{property}
describes a subset of
states.  In other words, a property is equivalent to a subset of the states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of properties.
Figure~\ref{fig:states} depicts a Venn diagram of various properties
and a trace.
An \emph{invariant}
\index{invariant}
is a property that holds over all reachable states.
We want to show that mutual exclusion is an invariant.
In other words, we want to show that the set of reachable states is a subset
of the set of states where there is at most one process in the critical
section.

One way to prove that a property is an invariant is through induction
on the number of steps.  First you prove that the property holds over
the initial state.  Then you prove that for every reachable state,
and for every step from that reachable state, the property also holds
over the resulting state.
For this you need a predicate that describes exactly which
states are reachable.
But we do not have such a predicate: we know how to describe the set
of reachable states, but given an arbitrary state it is not easy to
see whether it is reachable or not.

@@

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} mutual
exclusion.  In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

@@

To see that the inductive invariant implies mutual exclusion, suppose not.  Then
when both process~0 and process~1 are in the critical section, the
following must hold:
$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
 (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
We know that both flags are set.
We also know that neither process~0 nor process~1 is at label \texttt{@gate}
(because they are both at label \texttt{@cs}),
so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
logical impossibility.  Thus $I_p(i)$ implies mutual exclusion.

@@

So far we have pursued two properties: \emph{mutual exclusion}
and \emph{progress}.  The former is an example of a
\emph{safety property}---it prevents something ``bad'' from
happening, like a reader and writer process both entering the
critical section.  The \emph{progress} property is an example
of a \emph{liveness property}---guaranteeing that something good
eventually happens.
Informally (and inexactly), progress states that if no processes
are in the critical section, then some process that wants to enter
can.

>>>>

\newglossaryentry{atomicity}
{
  name=atomicity,
  description={describes that a certain machine instruction or sequence of machine
    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}
}
\newglossaryentry{concurrent execution}
{
  name=concurrent execution,
  description={(aka parallel execution) is when there are multiple processes executing and
      their machine instructions are interleaved in an unpredictable manner}
}
\newglossaryentry{context}
{
  name=context,
  description={(aka continuation) describes the state of a running process,
    including its program counter, the values of its variables (stored in its
    register), and the contents of its stack}
}
\newglossaryentry{critical section}
{
  name=critical section,
  description={(aka critical region) is a set of instructions that only one process
    is allowed to execute at a time.  The instructions are, however, not executed
    atomically, as other processes can continue to execute and access shared
    variables}
}
\newglossaryentry{determinism}
{
  name=determinism,
  description={is when the outcome of an execution is uniquely determined by the
    initial state}
}
\newglossaryentry{invariant}
{
  name=invariant,
  description={is a binary predicate over states that must hold for every
  reachable state of a process}
}
\newglossaryentry{machine instruction}
{
  name=machine instruction,
  description={is an atomic operation on the CXL machine, executed by a process}
}
\newglossaryentry{macro step}
{
  name=macro step,
  description={is a sequence of micro steps with at most one instruction
        externally visible to other processes.  It is the unit of interleaving
        in CXL}
}
\newglossaryentry{micro step}
{
  name=micro step,
  description={is the execution of one machine instruction by a process, updating
        its state and the state of the CXL machine}
}
\newglossaryentry{model checking}
{
  name=model checking,
  description={is a formal verification method that explores all possible executions
    of a program, which must have a finite number of states}
}
\newglossaryentry{process}
{
  name=process,
  description={is a method in execution.  We do not make the distinction
      between processes and threads.  A process has a program counter, a register,
      and a stack}
}
\newglossaryentry{process variable}
{
  name=process variable,
  description={is a variable that is private to a single process and stored in
        its register}
}
\newglossaryentry{property}
{
  name=property,
  description={describes a set of execution traces or histories that are allowed by
    a program.  Safety properties are properties in which ``no bad things happen,''
    such as violating mutual exclusion in a critical section.  Liveness properties are
    properties where ``something good eventually happens,'' like processes being
    able to enter the critical section if they want to}
}
\newglossaryentry{sequential execution}
{
  name=sequential execution,
  description={is when there is just one process executing, as opposed to concurrent
    execution}
}
\newglossaryentry{shared variable}
{
  name=shared variable,
  description={is a variable that is stored in the memory of the CXL machine and
      shared between multiple processes, as opposed to a process variable}
}
\newglossaryentry{stack machine}
{
  name=stack machine,
  description={is a model of computing where the state of a process is kept on
    a stack.  CXL uses a combination of a stack machine and a register-based machine}
}
\newglossaryentry{thread safety}
{
  name=thread safety,
  description={is when the implementation of a data structure allows concurrent access
    with well-defined semantics}
}

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
\glsadd{context}
it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
The CXL machine is predominantly a \emph{stack machine}.
\index{stack machine}
\glsadd{stack machine}
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process (with name tag \texttt{\_\_init\_\_/()}
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}:
\index{critical section}
\glsadd{critical section}
\index{critical region}
a set of instructions where only one process is allowed to execute at a
time.

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) is called \emph{thread-safe}.
\index{thread safety}
\glsadd{thread safety}
\index{thread}

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/states-crop.pdf}
\end{center}
\caption{Venn diagram classifying all states and a trace.}
\label{fig:states}
\end{figure}

@@

It is often useful to classify states into sets.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of classes of states.
Figure~\ref{fig:states} depicts a Venn diagram of various classes of states
and a trace.
One way to classify states is to define a predicate over states.
All states in which \texttt{x = 1}, or all states where
there are two or more processes executing, are examples of such predicates.
For our purposes, it is useful to define a predicate that says that at
most one process is in the critical section.  We shall call such states
\emph{exclusive}.

An \emph{invariant} of a program
\index{invariant}
is a predicate that holds over all states that are reachable by that program.
We want to show that exclusivity is an invariant.
In other words, we want to show that the set of reachable states of executing
the program
is a subset of the set of states where there is at most one process in the critical
section.

One way to prove that a predicate is an invariant is through induction
on the number of steps.  First you prove that the predicate holds over
the initial state.  Then you prove that for every reachable state,
and for every step from that reachable state, the predicate also holds
over the resulting state.
For this you need a predicate that describes exactly which
states are reachable.
But we do not have such a predicate: we know how to describe the set
of reachable states, but given an arbitrary state it is not easy to
see whether it is reachable or not.

@@

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

@@

To see that the inductive invariant implies exclusivity, suppose not.  Then
when both process~0 and process~1 are in the critical section, the
following must hold:
$(\lnot C(0) \lor \mathtt{process}(1)@gate) \land
 (\lnot C(1) \lor \mathtt{process}(0)@gate)$.
We know that both flags are set.
We also know that neither process~0 nor process~1 is at label \texttt{@gate}
(because they are both at label \texttt{@cs}),
so this simplifies to $(\lnot C(0)) \land (\lnot C(1))$.
So we conclude that $\mathtt{turn} = 0 \land \mathtt{turn} = 1$, a
logical impossibility.  Thus $I_p(i)$ implies exclusivity.

@@

A \emph{property}
\index{property}
\glsadd{property}
is a set of traces.
If a program has a certain property, that means that the traces that that
program allows are a subset of the traces in the property.
So far we have pursued two properties: \emph{mutual exclusion}
and \emph{progress}.  The former is an example of a
\emph{safety property}---it prevents something ``bad'' from
happening, like a reader and writer process both entering the
critical section.  The \emph{progress} property is an example
of a \emph{liveness property}---guaranteeing that something good
eventually happens.
Informally (and inexactly), progress states that if no processes
are in the critical section, then some process that wants to enter
can.
